---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
1.Load the data
```{r}
dat<- read.table(unz("activity.zip", "activity.csv"), header=T, quote="\"", sep=",")
```
2.Process the data into format suitable for analysis.
```{r}
dat$date<-as.Date(dat$date)
summary(dat)
```
## What is mean total number of steps taken per day?
1. Histogram of the total number of steps taken each day. 
```{r message=FALSE}
library(dplyr)
steps_per_day<-dat%>%
  group_by(date)%>%
  summarize(steps=sum(steps,na.rm=TRUE))
```

```{r}
hist(steps_per_day$steps,main="Distribution of Steps per day \n in October-November 2012 ",xlab="Total Steps Taken per day")

```


2. Calculate and report the mean and median total number of steps taken per day. 
```{r computecentre}
mean_steps<-round(mean(steps_per_day$steps),0)
median_steps<-median(steps_per_day$steps)
```
The mean steps per day is `r mean_steps`. The median steps per day is `r median_steps`.

## What is the average daily activity pattern?
1. Make a time series plot (i.e. type="l") of the 5-minute interval(x-axis) and the average number of steps taken, averaged across all days (y-axis). 
```{r message=FALSE}
steps_by_interval<-dat%>%
  group_by(interval)%>%
  summarize(steps=mean(steps,na.rm=TRUE))

par(mar=c(5,5,5,5),oma=c(1,1,2,1))
with(steps_by_interval,plot(x=interval, y=steps, type="l",main="Average Daily Activity Pattern \n in October to November 2012", xlab="Interval of the Day\n (5-minute increments)",ylab="Average number of \n steps taken"))
```

2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? 
```{r}
int_max_steps<-steps_by_interval[steps_by_interval$steps==max(steps_by_interval$steps),]
```
The `r int_max_steps[,1]`^th^ interval contains the maximum number of steps of `r round(int_max_steps[,2],0)`.

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset. 
```{r}
dat_na<-sum(is.na(dat))
```
The number of rows with missing values is `r dat_na`.

2. The strategy for filling in all of the missing values in the datase is to input the mean for the 5 minute interval. 
```{r}
steps_by_interval$steps<-round(steps_by_interval$steps,0)
```
3. Create a new dataset that is equal to the original dataset but with the missing data filled in. 
```{r}
dat_subset1<-dat[is.na(dat),]
dat_subset2<-na.omit(dat)
dat_subset3<-merge(dat_subset1,steps_by_interval,by="interval",all.x=TRUE)
dat_subset3<-dat_subset3[,-2]
names(dat_subset3)[3]<-"steps"
dat_new<-rbind(dat_subset3,dat_subset2)
```
4. Make a histogram of the total number of steps taken each day. 
```{r message=FALSE}
steps_per_day_new<-dat_new%>%
  group_by(date)%>%
  summarize(steps=sum(steps))

hist(steps_per_day_new$steps,main="Distribution of Steps per day \n in October-November 2012 ",xlab="Total Steps Taken per day")
```

Calculate and report the mean and median total number of steps taken per day. 
```{r}
mean_steps_new<-as.integer(mean(steps_per_day_new$steps),0)
median_steps_new<-as.integer(median(steps_per_day_new$steps))

```
After imputing the missing step values, the mean steps per day is `r mean_steps_new` and the median steps per day is `r median_steps_new`, Imputing the missing values has increased the estimates of the total number of steps. 

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels indicating whether a given date is a weekday or weekend day. 
```{r}
dat_new$weekday<-weekdays(dat_new$date)
dat_new<-mutate(dat_new, day = ifelse(weekday %in% 
                               c("Sunday","Saturday"),
                             "weekend","weekday"))
dat_new$day<-as.factor(dat_new$day)
            
```
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r message=FALSE}
steps_by_day<-dat_new%>%
  group_by(day,interval)%>%
  summarize(steps=mean(steps,na.rm=TRUE))

library(ggplot2)

g<-ggplot(steps_by_day,aes(interval,steps))
g+geom_line()+facet_grid(day ~ .)+
  xlab("Interval of the Day\n (5-minute increments)") + 
  ylab("Average number of \n steps taken") +
  ggtitle("Average Daily Activity Pattern \n in October to November 2012 \n by Day of the week")+
  theme(plot.title = element_text(hjust = 0.5))

```

